---
title: "Module 9 -- Exploratory Data Analysis & plotting"
author: "Loyal Goff"
date: "2022-11-02"
output:
  html_document:
    toc: true
    toc_level: 3
    toc_float: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Import

```{r}
# setwd("modules/module_9/notebooks")
library(SummarizedExperiment)
library(tidyverse)
library(dplyr)
# GEO <- "GSE63482"

# Read in the data
data <- read.csv("https://github.com/gofflab/Quant_mol_neuro_2022/raw/main/modules/module_9/notebooks/GSE63482_Expression_matrix.tsv", sep = " ", header = T, row.names = 1)
```

```{r}
# Parse header into parameters
sample_metadata <- as.data.frame(stringr::str_split_fixed(colnames(data), "_", 2))
colnames(sample_metadata) <- c("Age", "CellType")
sample_metadata$SampleID <- colnames(data)
colData <- DataFrame(sample_metadata)

# Create Gene metadata
gene_metadata <- data.frame(geneName = rownames(data))
rownames(gene_metadata) <- rownames(data)
gene_metadata <- DataFrame(gene_metadata)

# Create a Summarized Experiment object
se <- SummarizedExperiment(
  assays = SimpleList(fpkm = as.matrix(data)),
  rowData = gene_metadata,
  colData = sample_metadata
)
```

# Introduction to dplyr (tidyverse)

Many tasks in data analysis require exploration, manipulation and visualization.
Often before you perform any statistical analysis. The tidyverse package
[`dplyr`](https://dplyr.tidyverse.org/) ('data plier') was developed to help
with the common tasks manipulation, summarization, and organization of data.
Similar to the 'grammar of graphics' (ggplot2), `dplyr` attempts to define a
'grammar of data'. With this structure you can learn to quickly reshape, group,
summarise, and visualize your data to address your questions of interest.

## dplyr verbs

In the dplyr 'grammar of data', functions can be thought of as 'verbs' that
operate on a data.frame. Each verb performs a single function and does so
reasonably fast and efficiently. verb actions can be chained together to enable
powerful reorganization, summarizations, or manipulations of the input
data.frame.

_All dplyr functions take, as their first argument a data.frame object._ Let's
generate one for all of the gene expression data from our se object by making a
'long-form' melted and merged data.frame

```{r}
# this melt function will melt and merge an entire se object worth of data into a data.frame
meltSE <- function(se) {
  fpkm <- as.data.frame(assay(se, "fpkm"))
  fpkm$gene <- rownames(fpkm)
  fpkm.melt <- reshape2::melt(fpkm)
  colnames(fpkm.melt) <- c("gene", "SampleID", "value")
  tmp <- merge(fpkm.melt, colData(se), by.x = "SampleID", by.y = "SampleID")
  return(as.data.frame(tmp))
}

# Now lets create a melted and merged data.frame from our se object.
se.melt <- meltSE(se)

head(se.melt)
```

## dplyr verbs

- `select()` - selects columns of a data.frame
- `filter()` - selects rows of a data.frame matching a provided criterion
- `arrange()` - sorts a data.frame
- `mutate()` - adds columns to a data.frame based on a formula or equation
- `group_by()` - creates groups of rows that can be operated on collectively
- `summarise()` - creates new columns based on summarized properties of row
  groups

### Subsetting

The first two verbs define subsets of the data frame on either dimension.

```{r}
# select returns subsets of the columns in the order they are requested
select(se.melt, Age)
select(se.melt, Age, value)
select(se.melt, value, Age)
select(se.melt, -Age)

# filter() to select rows matching a logical equation
filter(se.melt, Age == "P1")
filter(se.melt, CellType %in% c("corticothal", "cpn"))
filter(se.melt, gene %in% c("Pax6", "Fezf2"))
filter(se.melt, Age != "P1")
filter(se.melt, value >= 5000)
filter(se.melt, Age == "P1" & CellType == "subcereb")
```

### Transforming

Often times we need to transform columns of the data (think of log-transforming
the 'values' column here as an example). This is performed with the `mutate()`
function in dplyr. And when we do this, we will need to provide a name for this
new column

```{r}
head(mutate(se.melt, logValue = log10(value + 1)))
```

Reminder that the output of dplyr can be assigned to a new variable as needed

```{r}
se.melt.transformed <- mutate(se.melt, logValue = log10(value + 1))
```

### Sorting

To sort a data.frame using dplyr we will use the `arrange()` function. This
takes (as always) a data.frame as the first argument, and then a list of columns
that you would like to sort in order.

```{r}
head(arrange(se.melt, Age))
head(arrange(se.melt, CellType))
head(arrange(se.melt, value))
head(arrange(se.melt, gene, Age, CellType))
```

### Chaining (piping) verbs together

dplyr provides (through the package `magrittr`) an additional way to 'chain'
verb actions together using a pipe operator (`%>%`). Conceptually, this is
equivalent to the Unix pipe operator (`|`) that we have seen previously. This
allows you to construct 'workflow' sentences in dplyr to perform actions in
order. In this context, it's important to remember that each dplyr function
takes a data.frame as it's first argument, and what the `%>%` is doing is
passing the output data.frame of the previous statement into the input argument
of the next.

```{r}
se.melt %>%
  filter(CellType == "cpn") %>%
  arrange(gene, Age) %>%
  head()
```

### Calculating summaries across variables

This is where the real power of dplyr begins to emerge. We can use the last two
verbs here, `group_by()` and `summarise()` to group rows together and perform
operations on each of the groups independently.

For example, if we wanted to summarise the average expression across all genes
for each of the celltypes we could do the following:

```{r}
se.melt %>%
  group_by(CellType) %>%
  summarise(mean = mean(value))
```

And by Age

```{r}
se.melt %>%
  group_by(Age) %>%
  summarise(mean = mean(value))
```

But more importantly, we can use combinations of variables to make more
insightful groupings...

```{r}
se.melt %>%
  group_by(gene, Age) %>%
  summarise(mean = mean(value), sd = sd(value)) %>%
  head()
```

Grouping also works with mutate() to preserve original data.frame structure

```{r}
se.melt %>%
  filter(gene == "Pax6") %>%
  group_by(CellType) %>%
  mutate(CellType_mean = mean(value))

# Make mean-centered gene expression values for each gene
se.melt %>%
  mutate(logFPKM = log10(value + 1)) %>%
  group_by(gene) %>%
  mutate(geneMean = mean(logFPKM), mean_centered = logFPKM - geneMean) %>%
  arrange(gene) %>%
  head(12)
```

Some of the problemset questions for this week will ask you to integrate some
dplyr manipulations with ggplot2 visualizations of summarised data.

# Session Information

```{r}
sessionInfo()
```
