---
title: "Getting Started Instructions (ME.440.825)"
author: "Loyal Goff"
date: '2022-08-05'
output: 
  html_document:
    toc: true
    toc_level: 3
    toc_float: true
    theme: yeti
---

# Introduction

# Learning Objectives

# R Notebooks

Notebooks are designed to facilitate 'literate analysis', which aims to make data analysis more interpretable and transparent. With notebooks, code is embedded with the narrative of the analysis. This has the benefit of helping to convey the specific ideas and thought processes of the author, and allows the introduction of concepts in the workflow that helps to facilitate understanding.

RStudio provides several tools to help develop and share R Notebooks, including the development and maintenance of a specific language structure called RMarkdown. RMarkdown combines the programming language R, with the simple markup language [Markdown](https://daringfireball.net/projects/markdown/). Markdown was designed to help convert human-readable text to HTML or PDF documents using a simple syntax.

-   [RMarkdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
-   [RMarkdown Video Overview](https://vimeo.com/178485416)

An R Markdown document begins with a short header section that describes the document that will be created. For example, this R Notebook has the following header at the top:

    ---
    title: "Intro to RStudo and RNotebooks"
    output: 
      html_notebook:
        toc: true
        toc_depth: 2
        toc_float: true
    editor_options: 
      chunk_output_type: inline
    ---

This header uses the [YAML](https://en.wikipedia.org/wiki/YAML) language syntax to define a few parameters for the document such as "title", "author", and "output". For the above header, we are letting RStudio know that we would like to output the final document as an `html_notebook`, with a table of contents (`toc: true`) and some other aesthetic elements.  

# RMarkdown Basics

# Chunks
The remainder of an Rmarkdown document is a combination of normal text (with markup for styling) and code. Code can be incorporated inline by wrapping it in a backticks (e.g. typing `` `r '\x60r pi\x60'` `` will add `r pi` inline with the rest of your text), or more often, by defining code chunks that contain both your code, and often instructions on how RStudio should evaluate/execute/display the results.


## Non-code chunks

In modern RMarkdown, 

## Code chunks

### Running a code chunk

```{r }
x<-rnorm(100)
mean(x)
```



# Knitting an R Notebook

# Rmarkdown cheat sheet
RStudio.com provides a useful [RMarkdown cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) which describes many tips, tricks, and syntax help for composing Rmarkdown documents.  You may find it handy to refer to this as you progress.

# Session Information
```{r session}
sessionInfo()
```
