## Part 1
```{r}
library(here)
tuesdata <- tidytuesdayR::tt_load("2020-01-21")
spotify_songs <- tuesdata$spotify_songs

# tests if a directory named "data" exists locally
if (!dir.exists(here::here("data"))) {
    dir.create(here::here("data"))
}

# saves data only once (not each time you knit a R Markdown)
if (!file.exists(here::here("data", "spotify_songs.RDS"))) {
    url_csv <- c(getwd(), "data/spotify_songs.csv")  # "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv"
    spotify_songs <- readr::read_csv(url_csv)

    # save the file to RDS objects
    saveRDS(spotify_songs, file = here("data", "spotify_songs.RDS"))
}

spotify_songs <- readRDS(here("data", "spotify_songs.RDS"))
df <- dplyr::as_tibble(spotify_songs)
```

```{r}
library(dplyr)
# How many songs are in each genre?
df %>%
    dplyr::count(playlist_genre) %>%
    dplyr::arrange(desc(n))
```

Note: I like to put the package name in front of the function.
This makes it easier to see where the function comes from.
It also makes it easier to see if you are using a function from a package that you haven't loaded yet.
It's best practices in every other programming languages.

```{r}
# What is average value of energy and acousticness in the latin genre in this dataset?
df %>%
    dplyr::filter(playlist_genre == "latin") %>%
    dplyr::summarise(mean_energy = mean(energy), mean_acousticness = mean(acousticness))
```

```{r}
# Calculate the average duration of song (in minutes) across all subgenres.
# Which subgenre has the longest song on average?
df %>%
    dplyr::group_by(playlist_subgenre) %>%
    dplyr::summarise(mean_duration = mean(duration_ms) / 60000) %>%
    dplyr::arrange(desc(mean_duration))
```

```{r}
# Make two boxplots side-by-side of the danceability of songs stratifying by whether a song has a fast or slow tempo.
# Define fast tempo as any song that has a tempo above its median value.
# On average, which songs are more danceable?
library(ggplot2)
df %>%
    dplyr::mutate(tempo = ifelse(tempo > median(tempo), "fast", "slow")) %>%
    ggplot(aes(x = tempo, y = danceability)) +
    geom_boxplot()
```

Note: modern practices discourage the use of bar graphs and boxplots when displaying distributional data
as they can be misleading for non-normal data.
Instead, use a violin plot or a [raincloud](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-and-how-to-create-them-with-ggplot2/) plot.

```{r}
to_plot <- df %>% dplyr::mutate(tempo = ifelse(tempo > median(tempo), "fast", "slow"))
ggplot(to_plot, aes(x = danceability, y = tempo, fill = tempo)) +
    geom_violin()
```

However, as we can see, this violin plot is not the easiest to read.
A ridgeline plot should be better.

```{r}
ggplot(to_plot, aes(x = danceability, y = tempo, fill = tempo)) +
    ggridges::stat_density_ridges(
        quantile_lines = TRUE,
        quantiles = 2,
        alpha = 0.7,
    ) +
    ggridges::theme_ridges() +
    labs(title = "Dancability. Line indicates median.")
```


## Part 2

```{r}
sce <- scRNAseq::MacoskoRetinaData()
# Show the number of number of genes and number of observations in the sce object.
dim(sce)
```

```{r}
# Quality control (using mitochondrial genes).
is.mito <- grepl("^MT-", rownames(sce))
qcstats <- scater::perCellQCMetrics(sce, subsets = list(Mito = is.mito))
filtered <- scater::quickPerCellQC(qcstats, percent_subsets = "subsets_Mito_percent")
sce <- sce[, !filtered$discard]
```
Try not to create a bunch of objects. They are duplicates and eat your memory.

```{r}
# Normalization.
sce <- scater::logNormCounts(sce)
```

`logNormCounts` outputs log-transformed normalized expression.
Normalized expression divides expression by the size factor for each cell,
computed as the library size of each cell after scaling them to have a mean of 1 across all cells.

**This is the simplest way for normalization. Big cells will have a higher size factor than small cells.
It also assumes that upregulation for a subset of genes is cancelled out by the same magnitude of downregulation in a different subset of genes.**

See [this](http://bioconductor.org/books/3.13/OSCA.basic/normalization.html).

```{r}
# Feature selection.
dec <- scran::modelGeneVar(sce)
hvg <- scran::getTopHVGs(dec, prop = 0.1)
```

> A larger subset will reduce the risk of discarding interesting biological signal by retaining more potentially relevant genes,
at the cost of increasing noise from irrelevant genes that might obscure said signal.
It is difficult to determine the optimal trade-off for any given application as noise in one context may be useful signal in another.

See [this](http://bioconductor.org/books/3.13/OSCA.basic/feature-selection.html).

```{r}
# PCA.
set.seed(1234)
sce <- scater::runPCA(sce, ncomponents = 25, subset_row = hvg)

# Clustering.
colLabels(sce) <- scran::clusterCells(sce,
    use.dimred = "PCA",
    BLUSPARAM = bluster::NNGraphParam(cluster.fun = "louvain")
)
```

```{r}
# Visualization.
sce <- scater::runUMAP(sce, dimred = "PCA")
scater::plotUMAP(sce, colour_by = "label")
```
