---
title: "Module 10 - Hierarchical & k-means clustering"
author: "Loyal Goff"
date: "2022-11-06"
output:
  html_document:
    toc: true
    toc_level: 3
    toc_float: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Import
```{r}
library(SummarizedExperiment)
library(tidyverse)

# Read in the data
data <- read.csv("https://github.com/gofflab/Quant_mol_neuro_2022/raw/main/modules/module_9/notebooks/GSE63482_Expression_matrix.tsv", sep = " ", header = T, row.names = 1)
```

```{r}
# Parse header into parameters
sample_metadata <- as.data.frame(stringr::str_split_fixed(colnames(data), "_", 2))
colnames(sample_metadata) <- c("Age", "CellType")
sample_metadata$SampleID <- colnames(data)
colData <- DataFrame(sample_metadata)

# Create Gene metadata
gene_metadata <- data.frame(geneName = rownames(data))
rownames(gene_metadata) <- rownames(data)
gene_metadata <- DataFrame(gene_metadata)

# Create a Summarized Experiment object
se <- SummarizedExperiment(
  assays = SimpleList(fpkm = as.matrix(data)),
  rowData = gene_metadata,
  colData = sample_metadata
)
assay(se,'logfpkm')<-log10(assay(se,'fpkm')+1)
```

# Clean/process data to cluster
```{r}
clustData<-assay(se,'logfpkm')
clustData<-clustData[rowMeans(clustData)>=1,]
```

# Hierarchical clustering
First we will need to compute a distance matrix of the pairwise distances across samples. To do this we will use the `dist()` function. `dist()` operates over rows, so if we want to calculate the distance between samples, we will have to transpose (`t()`) clustData.
```{r}
clustData.dist<-dist(t(clustData))
clustData.dist
```

```{r}
hclust(clustData.dist)

plot(hclust(clustData.dist),main="Euclidean:complete")
```


```{r}
clustData.dist<-dist(t(clustData),method="manhattan")

plot(hclust(clustData.dist),main="Manhattan:complete")
```

Partition by choosing k
```{r}
clustData.hclust<-hclust(clustData.dist)
plot(clustData.hclust)
clus3<-cutree(clustData.hclust,k=3)
clus3

```

Partition by fixing height
```{r}
clustData.hclust<-hclust(clustData.dist)
plot(clustData.hclust)
height<-1000
abline(h=height,lty="dashed",col="red")
clusH1000<-cutree(clustData.hclust,h=height)
clusH1000

```

# K-means
Let's try and identify groups of genes that behave similarly across the conditions. To do this, and to avoid the complication of absolute expression level (we're interested in 'patterns' across conditions), we will first center & scale the row(gene) data.
```{r}
nClusters<-12
clustData.scaled<-t(scale(t(clustData)))
clustData.kmeans<-kmeans(clustData.scaled,nClusters)
```

```{r}
clustData.plot<-clustData.scaled
clustData.plot<-as.data.frame(clustData.plot)
clustData.plot$gene<-rownames(clustData.plot)
clustData.plot$cluster<-clustData.kmeans$cluster
clustData.melt<-reshape2::melt(clustData.plot,id.vars=c("gene","cluster"))
clustData.melt<-merge(clustData.melt,colData(se),by.x="variable",by.y="SampleID")
clustData.melt<-as.data.frame(clustData.melt)
```

Summarise data across clusters with dplyr
```{r}
kmeans.summary <- clustData.melt %>%
  group_by(cluster,Age,CellType) %>%
  summarise(mean=mean(value),sd=sd(value))
```

```{r}
p<-ggplot(kmeans.summary,aes(x=Age,
                             y=mean,
                             ymax=mean+sd,
                             ymin=mean-sd
                             )) + 
  geom_point(aes(color=CellType)) +
  geom_line(aes(group=CellType,color=CellType)) +
  geom_hline(yintercept=0,linetype="dashed") + 
  #geom_ribbon(aes(group=CellType,color=CellType,fill=CellType),alpha = 0.2)
  theme_minimal()
p + facet_wrap('cluster',scales="free_y")
```

# Session Information
```{r}
sessionInfo()
```

