---
title: "Module 1 - Intro to R"
author: "Loyal Goff"
date: '2022-08-05'
output: 
  html_document:
    toc: true
    toc_level: 3
    toc_float: true
    theme: yeti
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
# Import libraries here 

```

# Introduction
<Overview of module topic>

# Learning Objectives
<List of learning objectives>

# What is R?

## Why R?
R is inherently a quirky language, but exceptionally powerful and well suited to data science, and bioinformatics in particular. R is not the fastest or most memory efficient programming language, but there are several benefits that make this a solid choice for learning genomics data science.

* R is free and open source and widely available across a variety of computational platforms
* R (& BioConductor) represent the efforts of an incredibly large, robust, and continually growing community dedicated to supporting and developing this language. 
* [BioConductor](https://www.bioconductor.org) is an incredibly rich and established ecosystem of data and code packages tailored specifically for a wide array of bioinformatics, genomics, and computational biology tasks. Chances are, if you're trying to solve a problem in computational biology, a package likely already exists to help you with this in Bioconductor.  We will use this resource widely in this course.
* R is one of the best languages for statistical modelling, and cleaning/organizing datasets. R has a strong foundation in statistical analysis and the syntax and community-based resources for performing many common statistical tasks are abundant.
* Visualzation of data in R is both beautiful and intuitive using packages such as `ggplot2` which define a 'grammar of graphics' that allows you to quickly build complex and stunning plots from your data.
* R can (seamlessly?) interact with other programming languages such as python or C++ within the same analysis.


## Rscripts

## RMarkdown documents

# The Basics
## Variables
R output can be created by directly typing equations into the console

```{r}
8 + 2
4**3
```
When executed, the above two lines simply return the result of the equations as output rows. But in order to use R as more than just a calculator, we need to be able to assign values to specific objects called variables.

A variable is in essence a 'named storage' contanier that programs can manipulate. Varibles in R can store a single atomic vector, a group of vectors, or a combination of different objects.
Variables can be assigned values using several different operators in R.

```{r}
# Assignment using the equal operator.
a = 5

# Assignment using the leftward operator (<-)
b <- 10

# Assignment using the rightward operator (->)
15 -> c
```
For historical reasons, the most common assignment operator in R is the leftward operator `<-`. We will continue to use this syntax for the remainder of the tutorial.

You will notice that in the above code chunk, no output was produced. By default, when a value is assigned to a variable, R does not print anything. You can inspect the contents of a variable by using the `print()` function or by typing the object name

```{r}
print(a)
b
c
```

## Data Types

## Number Types

## Strings

## Booleans

## Collection Data Types

### Vectors
The most basic data type in R is the vector, which consists of a group of values.  Vectors are created using the `c()` function as so:

```{r}
x <- c(5, 10, 15)
x
```

You can also store character strings in vectors as well. Individual elements of a string vector must be contained within quotes.

```{r}
y <- c("Toad", "the", "Wet", "Sprocket")
y
```

Many functions exist to either inspect or operate over vectors.
The `length()` function returns the length of the vector:
```{r}
length(x)
length(y)
```

```{r}
str(x)
str(y)
```

```{r}
class(x)
class(y)
```

What happens if we create a vector from two different types of variables, one numeric and one a character string?

```{r}
z <- c(1, "Radiohead", 4.0)
z
```
Elements within vectors in R **must** all be of the same data type. 

### DataFrames

## Control Flow

### if[else] and choice()

```
if (condition) true_action
if (condition) true_action else false_action
```

### Operators

#### Arithmetic Operators

#### Comparison Operators

#### Assignment Operators

### Looping/Iteration

#### For loops

#### While loops

####


# Session Information
```{r session}
sessionInfo()
```

